---
title: "Image Alignment in R"
editor_options: 
  chunk_output_type: inline
---

# Setup

```{r setup}
if(!requireNamespace("imager", quietly = T)) install.packages("imager")

library(imager)
library(rcell2.examples) # devtools::load_all()
library(tidyverse)
```

# Tests with imager

- https://cran.r-project.org/web/packages/imager/vignettes/gettingstarted.html
- http://dahtah.github.io/imager/
  - https://dahtah.github.io/imager/imager.html
  - https://dahtah.github.io/imager/morphology.html
  - https://github.com/ShotaOchi/imagerExtra
  - https://cran.r-project.org/web/packages/imagerExtra/vignettes/gettingstarted.html

## Methods

Functions defined in this package:

```{r}
?crop_cimg_around_xy
?fftshift
?cross_correlation
?img_alignment_offset
```

## Data

```{r}
data.dir <-
  "~/Projects/PhD/gitlabs_acl/rtcc/kar4/2024-08-22-RtCC-Ali-5-cepas-y-CFP/data/2024-08-22-RtCC-Ali-5-cepas-y-CFP/renamed/"
```


```{r}
# Load two images
frame1 <- file.path(data.dir, "BF_Position01_time02.tif") |> load.image()
frame2 <- file.path(data.dir, "BF_Position01_time04.tif") |> load.image()
```


```{r}
x <- 495
y <- 530
w <- 51

# Crop frame1.
img1 <- frame1 |> crop_cimg_around_xy(x, y, w) #|> plot()

# Crop frame2.
img2 <- frame2 |> crop_cimg_around_xy(x, y, w) #|> plot()

# Display the cropped images.
par(mfrow = c(1, 2))
plot(img1, main="Reference")
plot(img2, main="Query")
```

## Test

```{r}
cross_correlation(img1, img2) |> plot()
```


```{r}
# Calculate the offset through cross-correlation.
xy_offsets <- img_alignment_offset(img1, img2)

# Crop from the original image.
aligned_image <- crop_cimg_around_xy(frame2, x=x-xy_offsets[1], y=y-xy_offsets[2], w=w)

# Align the cropped image.
# aligned_image <- imager::imshift(img2, delta_x = xy_offsets[1], delta_y = xy_offsets[2])
```

```{r}
# Visualize the results
par(mfrow = c(2, 2))
cross_correlation(img1, img2) |> plot()
abline(h=xy_offsets["max_cor_y"], col="red")
abline(v=xy_offsets["max_cor_x"], col="red")
plot(img1, main = "Image 1")
plot(img2, main = "Image 2")
plot(aligned_image, main = "Image 2 (Aligned)")
```

# Test with EBIMage

No existe "computeTranslation", flasheó ChatGPT.

```{r}
# library(EBImage)
# 
# # Load images
# 
# # Load two images
# img1 <- file.path(data.dir, "BF_Position01_time02.tif") |> readImage()
# img2 <- file.path(data.dir, "BF_Position01_time04.tif") |> readImage()
# 
# # Compute translation using normalized cross-correlation
# translation <- computeTranslation(img1, img2)
# 
# # Apply the transformation
# aligned_img2 <- translate(img2, translation)
# 
# # Display the result
# display(combine(img1, aligned_img2))
```

# Tagging cells

The "multi-point" tool in ImageJ can be used to mark cells with different "counters",
by simply clicking on them in the image.

These marks can be exported and mapped back to `ucid`s in R, and then
used for filtering and analysis.

Steps:

1. Open the images in a Hyperstack, as described in the [Preview images using ImageJ] section.
2. Use the multi-point tool to mark individual cells with one or more "counters", as required by your experiment.
3. After tagging the cells, generate the measurements (press `Ctrl+M`) and save
   them to a `.csv` file (press `Ctrl+S`, or `File -> Save`).
  - It is critical that you check that the values of the "Ch" (channel), 
    "Slice", and "Frame" ImageJ measurements map correctly to `t.frame`s,
    `pos`, and `ch` in Cell-IDs output.
  - Note that ImageJ's time indexes begin at 1, but Cell-ID's 
    begin at 0. This may need to be adjusted later.

Detailed instructions on how to import these points and map them to cells 
identified by Cell-ID, can be found in the [Cell-tagging using ImageJ] section
of this notebook.

## Notas sobre el tagueo

1. Taguear celulas CFP en una cepa (6122, delta pho85).
  - Dio razonable. La mejor dosis para cuantificar entre cepas es 1 nM, porque se nota el efecto, y todas las cepas van a estar más a la derecha.
2. Taguear celulas CFP en la dosis de 1 nM, para comprobar que la dosis de feromona efectiva fue homogénea entre wells de la misma columna.
3. Taguear cepas del exp. a 1 nM.
  - *Todas* las células CFP en el well F6 (pos 37 y 38) terminaron **muertas**. No vi esto en el resto de las posiciones, pero tampoco lo busqué.
  - Se ve que las células de esas posiciones tienen una dinamica de switching diferente a la de las demas.

## Load tags

### Tag metadata

```{r}
counter_set1 <- bind_rows(
  data.frame(cell="cfp", Counter = c(0,1,4), tag = c("cell", "bud", "dead")),
  data.frame(cell="exp", Counter = 2:3, tag = c("cell", "bud"))
)

# counter_set1
```

### Control CFP Strain

```{r}
tags_dir <- 
  "~/Projects/PhD/gitlabs_acl/rtcc/kar4/2024-08-22-RtCC-Ali-5-cepas-y-CFP/data/tags"

tags_raw_cfp <- dir(tags_dir, pattern = "*.csv", full.names = T) |> 
  lapply(read.csv) |> 
  bind_rows() |>
  # filter(Counter %in% 1) |>
  left_join(counter_set1,by = join_by(Counter))

# tags_raw_cfp |> 
#   ggplot(aes(X, Y, color = as.factor(Counter))) +
#   geom_point() + 
#   facet_wrap(~Slice) +
#   theme_minimal() + 
#   scale_y_reverse()
```

```{r}
tags_cfp <- tags_raw_cfp |> 
  dplyr::rename(
    pos = Slice,
    t.frame = Frame
  ) |> 
  select(pos, t.frame, cell, tag, X, Y)

# tags_cfp
```

### Experiment Strains

```{r}
tags_raw_exp <- file.path(tags_dir, "exp/") |> dir(pattern = "*.csv", full.names = T) |> 
  lapply(read.csv) |> 
  bind_rows() |>
  filter(Counter %in% 2:3) |>
  left_join(counter_set1,by = join_by(Counter))

# tags_raw_exp |> 
#   ggplot(aes(X, Y, color = as.factor(Counter))) +
#   geom_point() + 
#   facet_wrap(~Slice) +
#   theme_minimal() + 
#   scale_y_reverse()
```


```{r}
tags_exp <- tags_raw_exp |> 
  dplyr::rename(
    pos = Slice,
    t.frame = Frame
  ) |> 
  select(pos, t.frame, cell, tag, X, Y)

# tags_exp
```

### Combine tags

```{r}
all_buds <- bind_rows(tags_cfp,tags_exp) |> 
  filter(tag == "bud")

all_buds
```


# Inspect cells

## Get image paths

```{r}
images <- rcell2.cellid::arguments(data.dir) |> 
  rcell2.cellid::arguments_to_images() |> 
  suppressWarnings()
# images$t.frame |> range()
```

## Generate cdata

```{r}
cdata <- all_buds |> 
  filter(t.frame > 3) |> 
  # mutate(t.frame = min(t.frame + 1, 65)) |> 
  dplyr::rename(xpos=X, ypos=Y) |> 
  mutate(ucid=1:n())
```

## Preview cells

```{r}
# rcell2.magick::magickCell(cdata, images, annotation_params = NULL,
#                           ch = "TFP", normalize_images = T, boxSize = 40, cell_resize = 100) |> 
#   rcell2.magick::magickForKnitr()
```

## Reframe cdata

Re-frame `cdata` to obtain rows for all frames for cell strips:

```{r}
# Transform each group to an arbitrary number of rows
cdata_adj <- cdata |> group_by(pos, ucid) |> 
  # Expand frames around the bud, from the beginning of the experiment.
  reframe(t.frame = 1:(t.frame+1)) |> 
  # Keep only valid frames that exist.
  semi_join(images) |> 
  # Keep only frames after the first, which has a focus shift.
  filter(t.frame > 1) |> 
  # Keep only sequences of sufficient length.
  group_by(ucid) |> 
  filter(n() >= 3) |> 
  # Reinsert the data.
  left_join(
    cdata |> rename(t.bud = t.frame)
  ) |> 
  # Remove cells with no frame after the bud.
  filter(t.bud < max(images$t.frame))

# cdata_adj |> 
#   ggplot() + geom_tile(aes(t.frame, ucid))
```

Check images:

```{r}
# cdata_adj |> 
#   rcell2.magick::cellStrips(images, n_ucids = 5, ch = "TFP",
#                             annotation_params = NULL,
#                             normalize_images = T, boxSize = 40, cell_resize = 100
#                             ) |> 
#   magick::image_join() |> 
#   magick::image_append(stack = T) |> 
#   rcell2.magick::magickForKnitr()
```

# Cell alignment

## Align two images

Sample:

```{r}
d_list <- cdata_adj |> split(~ucid)
# d <- d_list |> sample(1) |> first()
d <- d_list[[398]]

t.bud <- d$t.bud[1]

xy <- d |> filter(t.bud == t.frame) |> with(c(xpos, ypos))

img_paths <- images |> semi_join(d) |> 
  filter(channel == "TFP")
```

Process:

```{r}
# Set reference frame.
t.start <- t.bud
# Get query frame.
t.next <- d |> 
  filter(t.frame != t.start) |> 
  filter(t.frame == t.frame[ which.min(abs(t.frame - t.start)) ]) |> 
  with(t.frame)
```

```{r}
t.start <- 40
t.next <- 32
```

```{r}
# Get image paths.
path1 <- img_paths |> filter(t.frame == t.start) |> with(file)
path2 <- img_paths |> filter(t.frame == t.next) |> with(file)

# Load two images
frame1 <- load.image(path1)
frame2 <- load.image(path2)
```

Some unexpected stuff:

- Amazingly enough, 50 is the only window size that does not "drift" in X and Y by 1.
- I also have no idea why setting the widow size to odd numbers messes up the correlogram.

```{r}
# Set crop size, to not 50, and not odd.
w <- 52
# Set sigma.
s <- NULL
# Set blur sigma.
s2 <- NULL
```

```{r}
# Crop frame1.
img1 <- frame1 |> crop_cimg_around_xy(xy[1], xy[2], w)
# Crop frame2.
img2 <- frame2 |> crop_cimg_around_xy(xy[1], xy[2], w)

# Display the cropped images.
plot(img1, main="Reference")
abline(h=(w+1)/2, col="red")
abline(v=(w+1)/2, col="red")
plot(img2, main="Query")
abline(h=(w+1)/2, col="red")
abline(v=(w+1)/2, col="red")
```

Try smoothing the correlation signal a bit.

```{r}
par(mfrow = c(1, 3))

# Compute cross-correlation
cross_corr <- cross_correlation(img1, img2)
# Display the result
cross_corr |> as.cimg() |> plot(main = "Cross-correlation")

# Display blurred version.
cross_corr |> imager::isoblur(sigma=0.75) |> plot(main = "Iso-blur")

# Gaussian weights.
apply_gaussian_weighting(cross_corr[,,1,1], 250) |> as.cimg() |> plot(main = "Gaussian weighing")
```

```{r}
# Calculate the offset through cross-correlation.
xy_offsets <- img_alignment_offset(img1, img2, s=s, s2 = s2)

xy_offsets[c("offset_x", "offset_y")] <- xy_offsets[c("offset_x", "offset_y")] - 1

xy_offsets[c("offset_x", "offset_y")]
```

```{r}
# Crop from the original image.
aligned_image <- crop_cimg_around_xy(frame2, x=xy[1]-xy_offsets[1], y=xy[2]-xy_offsets[2], w=w)
```

```{r}
# Visualize the results
# par(mfrow = c(2, 2))
cross_correlation(img1, img2) |> plot()
abline(h=(w+1)/2, col="blue")
abline(v=(w+1)/2, col="blue")
abline(h=xy_offsets["max_cor_y"], col="red")
abline(v=xy_offsets["max_cor_x"], col="red")
plot(img1, main = "Image 1")
abline(h=(w+1)/2, col="red")
abline(v=(w+1)/2, col="red")
plot(img2, main = "Image 2")
abline(h=(w+1)/2, col="red")
abline(v=(w+1)/2, col="red")
plot(aligned_image, main = "Image 2 (Aligned)")
abline(h=(w+1)/2, col="red")
abline(v=(w+1)/2, col="red")
```

Note that pixel coordinates are not centered on the pixel in the plot below.

```{r}
m <- cross_correlation(img1, img2)

# m <- m |>
#   imager::isoblur(sigma=s2) |> 
#   as.matrix() |> 
#   apply_gaussian_weighting(s=s)

m_df <- m |> as.matrix() |> 
  t() |> 
  utiles::matrix_to_df() |> 
  arrange(-pix_value)

m_df_max <- m_df |> filter(pix_value == max(pix_value))

m_df |> 
  ggplot(aes(x,y)) +
  geom_tile(aes(fill=pix_value)) +
  geom_tile(fill="yellow", data=m_df_max) +
  geom_hline(yintercept=xy_offsets["max_cor_y"], alpha = .5, color="red") +
  geom_vline(xintercept=xy_offsets["max_cor_x"], alpha = .5, color="red") +
  geom_hline(yintercept=w/2, alpha = .5, linetype=2) +
  geom_vline(xintercept=w/2, alpha = .5, linetype=2) +
  coord_equal() + scale_y_reverse() + 
  ggtitle(paste("Max correlation at:", unlist(m_df_max)[1], unlist(m_df_max)[2], "(yellow)"), 
          paste("Final offsets at:", xy_offsets["max_cor_x"], xy_offsets["max_cor_y"], "(red)")) +
  theme_minimal()
```

## Methods

```{r}
?adjust_offset
?propagate_offsets
```

## Process a cell

Sample:

```{r}
d_list <- cdata_adj |> split(~ucid)

d <- d_list[[398]]
t.bud <- d$t.bud[1]

r <- 2:16
r <- 2:60
r <- 2:6
d <- d_list[[398]] |> 
  mutate(ypos = ypos + 30) |>
  filter(t.frame %in% r)
t.bud <- max(r)  # Reverse

xy <- d |>
  # filter(t.frame == 2) |> # Reverse
  filter(t.frame == max(r)) |> # Reverse
  with(c(xpos, ypos))

img_paths <- images |> 
  semi_join(d, by = join_by(pos, t.frame)) |> 
  filter(channel == "TFP") |> 
  arrange(t.frame)

pics <- d |> 
  rcell2.magick::magickCell(images, normalize_images = T, return_single_imgs = T, annotation_params = NULL,
                            n.cells=nrow(d), 
                            highlight_frames=t.bud,
                            ch = "TFP", boxSize=50) |> 
  rcell2.magick::square_tile(nRow = 3) |> 
  rcell2.magick::magickForKnitr()
```

Test:

```{r}
img_offsets <- img_paths |> 
  arrange(-t.frame) |> # Reverse.
  
  # Align.
  propagate_offsets(t.bud = t.bud, xy = xy, verbose = F) |> 
  
  # Compute cumulative offset.  
  mutate(x_offset_sum = cumsum(x_offset),
         y_offset_sum = cumsum(y_offset)) |> 
  select(pos, t.frame, 
         x_offset, y_offset, 
         x_offset_sum, y_offset_sum, t.ref)

# img_offsets
```

### BUG: exchanged offsets

I need to exchange the X and Y offset sums for alignment to work (or at least display) properly.

I have reviewed this extensively and haven't found the issue. whatever works...

```{r}
d_xy <- d |> 
  left_join(img_offsets) 

pics <- d_xy |> 
  
  # BUG HERE !!!!!
  mutate(xpos = xpos - y_offset_sum) |> # BUG HERE !!!!!
  mutate(ypos = ypos - x_offset_sum) |> # BUG HERE !!!!! 
  # BUG HERE !!!!!
  
  rcell2.magick::magickCell(images, normalize_images = T, annotation_params = NULL,
                            return_single_imgs = T,
                            n.cells=nrow(d), 
                            highlight_frames=t.bud,
                            ch = "TFP", boxSize=50)

# Preview.
pics |> 
  rcell2.magick::square_tile(nRow = 2) |>
  # magick::image_append() |>
  rcell2.magick::magickForKnitr()
```

```{r}
d_xy_img <- d_xy |>  
  select(pos, t.frame, xpos, ypos) |> 
  left_join(img_paths) |> 
  arrange(t.frame) |> 
  split(~t.frame) |> 
  lapply(function(d_corr){
    img <- load.image(d_corr$file)
    crop_cimg_around_xy(img, x = d_corr$xpos, y = d_corr$ypos, w = w)
  })

# Create a 2D tile with configurable rows and columns
tile <- create_image_tile(d_xy_img, n_rows = 5, aspect_ratio = 1)
plot(tile)
```


```{r}
d_xy_img <- d_xy |>  
  mutate(xpos = xpos - y_offset_sum) |>
  mutate(ypos = ypos - x_offset_sum) |> 
  select(pos, t.frame, xpos, ypos) |> 
  left_join(img_paths) |> 
  arrange(t.frame) |> 
  split(~t.frame) |> 
  lapply(function(d_corr){
    img <- load.image(d_corr$file)
    crop_cimg_around_xy(img, x = d_corr$xpos, y = d_corr$ypos, w = w)
  })

# Create a 2D tile with configurable rows and columns
tile <- create_image_tile(d_xy_img, n_rows = 5, aspect_ratio = 1)
plot(tile)
```

### Full time-course test

```{r}
d_list <- cdata_adj |> split(~ucid)

d <- d_list[[398]]

t.bud <- d$t.bud[1]

xy <- d |>
  filter(t.frame == t.bud) |>
  with(c(xpos, ypos))

img_paths <- images |> 
  semi_join(d, by = join_by(pos, t.frame)) |> 
  filter(channel == "TFP") |> 
  arrange(t.frame)
```

Calculate offsets:

```{r}
w <- 50
# s <- 250
# s2 <- 0.75

# w <- 25
s <- NULL
s2 <- NULL

off_drift_x <- 0 # 1
off_drift_y <- 0 # 1.5
```

```{r}
# After budding.
img_paths_post <- img_paths |> 
  filter(t.frame >= t.bud) |> 
  propagate_offsets(t.bud, xy, w, s, s2, xy_adj=F) |> 
  mutate(x_offset_sum = cumsum(x_offset-off_drift_x),
         y_offset_sum = cumsum(y_offset-off_drift_y))

# Before budding.
img_paths_pre <- img_paths |> 
  filter(t.frame <= t.bud) |> 
  arrange(-t.frame) |> 
  propagate_offsets(t.bud, xy, w, s, s2, xy_adj=F)|> 
  mutate(x_offset_sum = cumsum(x_offset-off_drift_x),
         y_offset_sum = cumsum(y_offset-off_drift_y))

# Join results.
img_offsets <- bind_rows(img_paths_pre, 
                         img_paths_post[-1,]) |> 
  arrange(t.frame) |> 
  select(pos, t.frame, 
         x_offset, y_offset, 
         x_offset_sum, y_offset_sum, t.ref)
# img_offsets
```

Mitigate "the BUG" here:

```{r}
# Generate images.
pics <- d |> 
  # left_join(img_paths_pre) |> filter(!is.na(file)) |> 
  left_join(img_offsets, by = join_by(pos, t.frame)) |>
  
  # BUG HERE !!!!!
  mutate(xpos = xpos - y_offset_sum) |> # BUG HERE !!!!!
  mutate(ypos = ypos - x_offset_sum) |> # BUG HERE !!!!! 
  # BUG HERE !!!!!
  
  rcell2.magick::magickCell(images, normalize_images = T, return_single_imgs = T, annotation_params = NULL,
                            n.cells=nrow(d),
                            highlight_frames=t.bud,
                            ch = "TFP", boxSize=50)

# Preview.
pics |> 
  rcell2.magick::square_tile(nRow = 5) |> 
  # magick::image_append() |> 
  rcell2.magick::magickForKnitr()
```

## Process all

```{r}
w <- 100
s <- 500
s2 <- 0.75

off_drift_x <- 1
off_drift_y <- 1.5
```

```{r}
w <- 50
s <- NULL
s2 <- NULL

off_drift_x <- 0
off_drift_y <- 0
```

```{r}
process_one <- function(d){
    t.bud <- d$t.bud[1]
  
    xy <- d |> filter(t.bud == t.frame) |> with(c(xpos, ypos))
    
    img_paths <- images |> 
      semi_join(d, by = join_by(pos, t.frame)) |> 
      filter(channel == "TFP") |> 
      arrange(t.frame)
    
    # After budding.
    img_paths_post <- img_paths |> 
      filter(t.frame >= t.bud) |> 
      propagate_offsets(t.bud, xy, w, s, s2) |> 
      mutate(x_offset_sum = cumsum(x_offset-off_drift_x),
             y_offset_sum = cumsum(y_offset-off_drift_y))
    
    
    # Before budding.
    img_paths_pre <- img_paths |> 
      filter(t.frame <= t.bud) |> 
      arrange(-t.frame) |> 
      propagate_offsets(t.bud, xy, w, s, s2)|> 
      mutate(x_offset_sum = cumsum(x_offset-off_drift_x),
             y_offset_sum = cumsum(y_offset-off_drift_y))
    
    # Join results.
    img_offsets <- bind_rows(img_paths_pre, 
                             img_paths_post[-1,]) |> 
      arrange(t.frame) |> 
      select(pos, t.frame, x_offset, y_offset, x_offset_sum, y_offset_sum, t.ref)
    
    # Make images.
    pics <- d |> 
      left_join(img_offsets, by = join_by(pos, t.frame)) |> 
  
      # BUG HERE !!!!!
      mutate(xpos = xpos - y_offset_sum) |> # BUG HERE !!!!!
      mutate(ypos = ypos - x_offset_sum) |> # BUG HERE !!!!! 
      # BUG HERE !!!!!
      
      rcell2.magick::magickCell(images,
                                normalize_images = TRUE, return_single_imgs = TRUE, 
                                annotation_params = NULL, add_border = FALSE,
                                ch = "TFP", boxSize=w, n.cells = nrow(d))
    
    # Write images.    
    for(j in 1:nrow(d)){
      p <- file.path(out.dir, paste0("cell-", d$ucid[1], "-time-", d$t.frame[j], "-bud-", t.bud, ".tiff"))
      magick::image_write(image = pics[j], path = p, format = "tiff")
    }
    
    return(pics)
  }
```


```{r}
out.dir <- file.path("/tmp", "cells")
dir.create(out.dir, F, T)

d_list <- cdata_adj |> split(~ucid)
# d_list <- d_list[1:5]

# d <- d_list[[398]]
r <- d_list |> 
  lapply(function(d){
    tryCatch(process_one(d),
             error = function(e) cat("Failed to process cell", d$ucid[1], "\n"))
  })
```


```{r}
cdata_adj |> 
  filter(t.frame == t.bud) |> 
  select(ucid, t.frame) |> 
  write.csv(file = "/tmp/cells.csv")
```

